---
title: "Introduction to dplyr"
author: "Fabrice Rossi"
format: 
  revealjs:
    echo: false
  pptx:
    echo: false
lightbox: true
df-print: kable
---

```{r}
here::i_am("dplyr-101.Rproj")
library(here)
```

```{r}
library(ggplot2)
theme_set(theme_bw())
library(dplyr)
```

```{r}
data("diamonds")
```

# Introduction to `filter`

## Link between diamond carat and price

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

## Selecting a subset of a dataset

```{r}
#| echo: true
diamonds_2ct <- diamonds |>
  filter(abs(carat - 2) <= 0.1)
```

```{r}
ggplot(diamonds_2ct, aes(x = carat, y = price)) +
  geom_point()
```

```{r}
nb_diamonds_near_2 <- diamonds_2ct |> nrow()
nb_diamonds_below_2 <- diamonds_2ct |>
  filter(carat < 2) |> 
  nrow()
```

## Gap in the distribution

There are:

- `r nb_diamonds_below_2` diamonds with weight between 1.9 and 1.99 carats
- `r nb_diamonds_near_2-nb_diamonds_below_2` diamonds with weight between 2 and 2.1 carats

There is a clear gap in the distribution of the weights, with a very small number of diamonds just before 2 carats compared to the number of them just after 2 carats.

## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x = carat, y = price, group = carat)) +
  geom_boxplot()
```

## Prices of 2 carat diamonds

```{r}
diamonds |>
  filter(carat >= 2, carat <= 2.1) |>
  ggplot(aes(x = carat, y = price, group = carat)) +
  geom_violin()
```

## Using filter {.smaller}

Filter is used as follows

```{r}
#| eval: false
#| echo: true
filter(dataframe, expression with a logical value)
```

Interpretation: the code keeps the rows of the data frame for which the expression evaluates to `TRUE`.

```{r}
filter(diamonds, price == 1000) |>
  ggplot(aes(y=cut)) +
    geom_bar()
```

## Multiple logical expressions {.smaller}

::::: columns
::: {.column width="40%"}
Filter can be used as follows

```{r}
#| eval: false
#| echo: true
filter(dataframe, exp1, exp2, exp3)
```

Interpretation: the code keeps the rows of the data frame for which all expressions evaluate to `TRUE`.
:::

::: {.column width="60%"}
```{r}
filter(diamonds, price == 1000, cut == "Ideal" ) |>
  ggplot(aes(y=color)) +
    geom_bar()
```
:::
:::::

# Logical expressions

## Comparison operators

Everything that compares two values and returns `TRUE` or `FALSE`.

```{r}
#| echo: true
x <- 2
x == 2
x >= 3
```

List: `<`, `<=`, `==`, `>=`, `>` and `!=`

## Logical operators

Combining true value.

```{r}
#| echo: true
TRUE & FALSE
TRUE | FALSE
!FALSE
```

- `|` logical OR
- `&` logical AND
- `!` logical NOT

## Examples

Two equivalent filter operations:

```{r}
#| echo: true
#| eval: false
diamonds |> filter(price == 1000, cut == "Ideal") 
diamonds |> filter( (price == 1000) & (cut == "Ideal") ) 
```

Two different filter operations:

```{r}
#| echo: true
#| eval: false
diamonds |> filter(price == 1000, cut == "Ideal") 
diamonds |> filter( (price == 1000) | (cut == "Ideal") ) 
```

# Introduction to `summarise`

## Global summaries

```{r}
#| echo: true
diamonds |> 
  summarise(mean(price), median(price))
```

`summarise` is used as follows:

```{r}
#| eval: false
#| echo: true
summarise(dataframe, expression based on an aggregation function)
```

The result is a data frame with one column per expression and one row.

## Aggregation functions {.smaller}

::::: columns
::: {.column width="50%"}
An aggregation function computes a statistics (in a broad sense) on a numerical vector, for instance

- `mean`
- `median`
- `sd`: standard deviation
- `var`: variance
- `max` and `min`
:::

::: {.column width="50%"}
```{r}
#| echo: true
x <- c(-0.1, 1, 2, 2, -3, 5)
mean(x)
median(x)
sd(x)
min(x)
n_distinct(x)
length(x)
```
:::
:::::

## Limitations of `summarise` {.smaller}

Aggregation functions are mostly limited to numerical values (and to logical ones to some extent). Aggregation functions for non numerical vectors generally output multiple values.

```{r}
#| echo: true
y <- c("A", "A", "B", "C", "A", "B", "C", "C", "D")
table(y)
```

```{r}
#| echo: true
diamonds |> 
  pull(cut) |>
  table()
```

This generates an error because the aggregation contains multiple values.

```{r}
#| echo: true
#| eval: false
diamonds |>
  summarise(table(cut))
```

## Renaming the resulting columns

```{r}
#| echo: true
diamonds |> 
  summarise(mean_price = mean(price), 
            median_price = median(price), 
            interquartile_price = IQR(price))
```

In many dplyr verbs, one can give to an expression a name in the form `name=expression`. The name will be used for the corresponding column.

# Grouped operations

## Grouped summaries {.smaller}

::::: columns
::: {.column width="50%"}
```{r}
#| echo: true
diamonds |>
  summarise(median_price = median(price), 
            .by = cut)
```
:::

::: {.column width="50%"}
Effects of `.by`

1)  split the data frame into as many sub-data frames as values taken by the `.by` variable
2)  apply the aggregation functions to each sub-data frame
3)  glue together the resulting one row data frames
4)  add a column to identify the subsets
:::
:::::

## Manual calculation for one value

```{r}
#| echo: true
diamonds |>
  filter(cut == "Ideal") |>
  summarise(median_price = median(price))
```

## Graphical investigation

```{r}
ggplot(diamonds, aes(x=price, colour = cut)) +
  geom_density(bw="sj", linewidth = 1, alpha = 0.75)
```

## Additional statistics

Statistics summarising the price distribution given the cut.

```{r}
diamonds |>
  summarise(median = median(price), 
            mean = mean(price),
            sd = sd(price),
            IQR = IQR(price),
            .by = cut) |>
  arrange(cut)
```

## Counting stuff

::::: columns
::: {.column width="50%"}
```{r}
#| echo: true
diamonds |> 
  summarise(count = n(), 
            .by = cut)
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
diamonds |> 
  group_by(cut) |> 
  summarise(count = n())
```
:::
:::::

## Dependencies

A standard way to assess dependency between two numerical variables is Pearson's correlation coefficient: `cor`.

::::: columns
::: {.column width="50%"}

```{r}
diamonds |>
  summarise(cor(carat, price))
```
:::

::: {.column width="50%"}
```{r}
diamonds |>
  summarise(cor(carat, price), .by=cut)
```
:::
:::::




## Dependencies (non numerical) {.smaller}

```{r}
diamonds |>
  ggplot(aes(x=cut, fill=color)) +
  geom_bar(position="stack")
```

## Dependencies (non numerical)

```{r}
diamonds |>
  summarise(n(), .by=c(cut, color)) |>
  slice(1:5) ## first five rows
```

# Operating on columns

## Introduction to `mutate`

The `mutate` verb is used to create new columns (or to modify existing ones)
based on calculation conducted on existing columns. Generic form:

```{r}
#| echo: true
#| eval: false
mutate(dataframe, newcolumn = expression)
```


```{r}
#| echo: true
diamonds_ppc <- diamonds |>
  mutate("price per carat" = price/carat)
```

## Application example

```{r}
diamonds_ppc |>
  summarise(median(`price per carat`), .by=cut) |>
  arrange(cut)
```

## Application example (graphical)

```{r}
diamonds_ppc |>
  ggplot(aes(x=`price per carat`, colour = cut)) +
  geom_density(bw="sj", linewidth = 1, alpha = 0.75)
```

